Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-01 paper/main.tex
Updates

While there has been some effort in
DAG literature as well towards allowing for expert knowledge but these methods
focus towards providing the expert knowledge to the automated algorithms
\citep{Bang2023, Meek1995}.

@@

Fig.~\ref{fig:intro} highlights some of these issues using the Adult income
dataset \citep{Becker1996}. As can be seen in the figure depending on the
sample size and the algorithm used, we get vastly different results.

As a result of these challenges users have relied on model testing approaches
along with their expert knowledge to construct these models. The other ways is
to do post hoc modification to the learned model using implied CI based model
testing \citep{Ankan2021, Ankan2023}. In this testing procedure, users test
whether the implied CIs of the DAG hold in the dataset or not using statistical
CI tests. Based on the results of these statistical tests, we can accordingly
make modifications to the current model to fix the violations. However, this CI
based testing also has a few drawbacks: 1) CIs are implied only by missing
edges in the model, so we can not test for already existing edges. 2) In high
sample size scenarios, p-values can suggest a significant association between
variables even if they have a very weak relationship. 3) Number of implied CIs
can be large making it difficult to manually go through each of them; for
example, in the moderately sized alarm network with $ 37 $ node, there are
$287$ implied CIs.

There has been work on integrating expert knowledge into model building process
and has been quite successful such as tpc and background knowledge
specification. There has been work on trying to come up with a single test to
combine all these test to give a single answer \citep{Eulig2023, Shipley2000}.
However, what is missing is the guide on how to how to fix these tests.
Additionally there has been a lot of work to integrate expert knowledge into
causal discovery algorithms which has shown significant improvement. To tackle
these issues with CI based testing, we take inspiration from \emph{modification
indices} in specification search. The modification indices method tries to
identify and rank changes in the model such that modifications that improve the
model fit the most are ranked higher.This ranking allows researchers to focus
first on the suggestions that can improve the model the most and while
integrating their expertise to choose the most optimal modifications. Analogous
to this, in this paper we develop a measure for partial association for mixed
data that allows us to rank the CI violations. Additionally, this measure of
conditional association also allows us tackle other issues with CI based
testing outlined in the previous paragraphs.

\begin{figure}
\begin{subfigure}{0.25 \textwidth}
	\includegraphics[page=1]{figures.pdf}
	\caption{PC algorithm $ \chi^2 $ test. $ N=400$ }
\end{subfigure}%
\begin{subfigure}{0.25 \textwidth}
	\includegraphics[page=2]{figures.pdf}
	\caption{PC algorithm $ \chi^2 $ test. $ N=800$ }
\end{subfigure}
\begin{subfigure}{0.25 \textwidth}
	\includegraphics[page=5]{figures.pdf}
	\caption{Hill Climb Search BIC Score. $ N=400$ }
\end{subfigure}%
\begin{subfigure}{0.25\textwidth}
	\includegraphics[page=6]{figures.pdf}
	\caption{Hill Climb Search BIC Score. $ N=800$ }
\end{subfigure}
\begin{subfigure}{0.25 \textwidth}
	\includegraphics[page=3]{figures.pdf}
	\caption{PC algorithm residualization test. $ N=400$ }
\end{subfigure}%
\begin{subfigure}{0.25\textwidth}
	\includegraphics[page=4]{figures.pdf}
	\caption{PC algorithm residualization test. $ N=800$ }
\end{subfigure}

\caption{Markov Equivalence Class (MEC) learned for Adult Income Dataset using (a) PC algorithm with chi-squared test, (b) Hill-Climb Search with BIC score, (c) PC algorithm with residualization based test. There is significant variation in the learned model depending on the sample size and the algorithm used. Additionally, all of them make some mistakes.}
\label{fig:intro}
\end{figure}

>>>>

@@

Fig.~\ref{fig:intro} highlights some of these issues. To deal with these issues
methods for integrating expert knowledge have been developed. The two main
methods to integrate expert knowledge in the model construction process is to:
1. Provide all the expert knowledge to the causal discovery algorithm for
example in the form of required and forbidden edges \citep{Meek1995}, or
temporal information \citep{Bang2023} and let the algorithm use it to construct
the model. 2. Utilize conditional independence (CI) testing to identify issues in
the model, and manually fix them based on expert knowledge \citep{Ankan2021}.
We focus on this second approach in this paper.

In this CI testing procedure, users test whether the implied CIs of the DAG
hold in the dataset or not using statistical CI tests. Based on the results of
these statistical tests, we can accordingly make modifications to the current
model to fix the violations. However, this CI based testing also has a few
drawbacks: 1) CIs are implied only by missing edges in the model, so we can not
test for already existing edges. 2) In high sample size scenarios, p-values can
suggest a significant association between variables even if they have a very
weak relationship. 3) Number of implied CIs can be large making it difficult to
manually go through each of them; for example, in the moderately sized alarm
network with $ 37 $ node, there are $287$ implied CIs.

There has been work on trying to come up with a single test to combine all
these test to give a single answer \citep{Eulig2023, Shipley2000}. However,
what is missing is the guide on how to how to fix these tests. Additionally
there has been a lot of work to integrate expert knowledge into causal
discovery algorithms which has shown significant improvement. To tackle these
issues with CI based testing, we take inspiration from \emph{modification
indices} in specification search. The modification indices method tries to
identify and rank changes in the model such that modifications that improve the
model fit the most are ranked higher.This ranking allows researchers to focus
first on the suggestions that can improve the model the most and while
integrating their expertise to choose the most optimal modifications. Analogous
to this, in this paper we develop a measure for partial association for mixed
data that allows us to rank the CI violations. Additionally, this measure of
conditional association also allows us tackle other issues with CI based
testing outlined in the previous paragraphs.

\begin{figure}
\begin{subfigure}{0.25 \textwidth}
	\includegraphics[page=1]{figures.pdf}
	\caption{$ N = 400 $}
\end{subfigure}%
\begin{subfigure}{0.25 \textwidth}
	\includegraphics[page=2]{figures.pdf}
	\caption{$ N = 800$ }
\end{subfigure}
\begin{subfigure}{0.25 \textwidth}
	\includegraphics[page=5]{figures.pdf}
	\caption{$ N = 400 $ }
\end{subfigure}%
\begin{subfigure}{0.25\textwidth}
	\includegraphics[page=6]{figures.pdf}
	\caption{ $ N = 800$ }
\end{subfigure}
\begin{subfigure}{0.25 \textwidth}
	\includegraphics[page=3]{figures.pdf}
	\caption{ $ N = 400$ }
\end{subfigure}%
\begin{subfigure}{0.25\textwidth}
	\includegraphics[page=4]{figures.pdf}
	\caption{$ N = 800$ }
\end{subfigure}

\caption{A comparison of Markov Equivalence Classes (MEC) learned using Adult Income Dataset \citep{Becker1996} using (a \& b) PC algorithm using a mutual information based test, (c \& d) Hill Climb Search with Bayesian Information Criterion (BIC) score, (e \& f) PC algorithm with a residualization based test \citep{Ankan2023}. The learned model structure varies significantly depending on the algorithm, the CI test, and the sample size used.}
\label{fig:intro}
\end{figure}
